---
title: "Analytics Using R - Contents"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Analytics Using R - `r Sys.Date()` </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Programme

 - 10:30 Start
 - 10:40 Use case 1 
 - 11:00 Hands-on Part 1; RStudio and R-basics
 - 11:45 -- 12:30 LUNCH
 - 12:30 Hands-on Part 2  
 - 15:00 Use case 2
 - 15:30 Hands-on Part 3
 - 17:00 Drinks & Networking

# Workshop contents

## **PART 1**

 * Start; "Reproducible research with R" - 'Jan-Willem Lankhaar'
 * Use case 1; "Big Data in Psychiatry" - 'Karin Hagoort'
 * Hands-on 1; Github, RStudio, R-basics
 
 ------ exercise ------

## **PART 2** 
 
 * Recap R-basics 
 * Hands-on 2; Visualizations with `{ggplot2}`
 
 ------ exercise ------
 
 * Use case "Open Analytics" - 30 min.
 * Hands-on 3; Data wrangling with `{dplyr}`
 
## Datasets

 * `gapminder`  
 * `heights_outliers`
 * `Animals.xlsx`  
 * `pertussis_outbreaks.csv`

# Credits

Much of this material has been derived in one way or the other from the work of 
Hadley Wickham and Garret Grolemund and many other. For a more elborate reference list see the REFERENCES.txt file in the project root.

Thanks to Hadley & Garret for writing the book "R for Data Science" http://r4ds.had.co.nz/ and for their work in general to innovate the R world.

Work on integration of Git/Github with R/RStudio is thouroughly and witfully documented by Jenny Brian. I also very much appreciate her work on the use of R-Markdown. See also:

https://github.com/jennybc/happy-git-with-r & http://stat545.com/block007_first-use-rmarkdown.html 

## Disclaimer & Licence

_The work in this presentation has been adapted from many sources as mentioned in slide 'Credits'. For a full reference list see the 'REFERENCES.txt' document in the github repo 
https://www.github.com/usashogeschoolutrecht/surf_research_bootcamp_2018

The work presented here may be shared, remixed or adapted as long as the original references and the authors of this document are mentioned in the redistibution: CC BY-SA



