---
title: 'Hands-on 3: Wrangling data with the {dplyr} package'
author: "Marc A.T. Teunis"
date: "11/2/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Citations
```{r}
citation(package = "ggplot2")
citation(package = "tidyverse")
citation(package = "dslabs")
```

## Packages
```{r}
library(tidyverse)
library(dslabs)
```

## Data

Pertussis outbreaks from World Health Organization

http://data.euro.who.int/cisid/?TabID=463987

See for info also: 

http://ecdc.europa.eu/sites/portal/files/documents/Pertussis%20AER.pdf

## Source data
The data used in this presentation has been constructed from the interactive datbase tool hosted at: http://data.euro.who.int/cisid/ 
A selection was made for all available countries and all avaialable years, for the number os cases reported to the WHO for Whooping Cough - pertussis infections.
The file is avaialble in `./data/CISID_pertussis_10082018.csv`

## Characteristics

- The first few lines of the file have comments indicated with `#`
- There are 53 countries in the dataset
- There is no data for the years 1980-1989 and 2014-2018
- The data is not in `tidy` format
- Missing values are indicated with _`NA`_
- It is not a good idea to have a column name starting with a digit; why?

## Inspect the file with a linux command
```{r, linux_command}
system("head -20 ./data/CISID_pertussis_10082018.csv")
```
The delimeter is `,` and we indeed see the comments at the start of the file. 

## Read data
```{r, read_data}
pertussis_data <- read_csv(file = file.path(root, 
                                            "data", 
                                            "CISID_pertussis_10082018.csv"),
                           comment = "#", n_max = 53, na = c("", " "))
```

## Inspect data
```{r, inspect_data}
pertussis_data
names(pertussis_data)
```

## The pipe

**%>%**

## Tidy data
```{r, eval=FALSE, error_variable_reference}

## try the code below, why does it fail?
pertussis_data_tidy <- pertussis_data %>% 
  gather(1980:2018, key = "year", value = "annual_pertussis_cases") %>%
  print()

```

Using ``` ` ``` to reference names that contain 'digits' or other 'special characters'
```{r, good_variable_reference}

pertussis_data_tidy <- pertussis_data %>% 
  gather(`1980`:`2018`, key = "year", value = "annual_pertussis_cases") %>%
  ## rename X1 / x2 columns
  rename(key = X1,
         country = X2) %>%
  mutate(annual_pertussis_cases = as.integer(annual_pertussis_cases)) %>%
# we prentend that the measurement for the year was reported on January 1st the next  year.
  mutate(year = lubridate::ymd(year, truncated = 2L)) %>%
  mutate(year = lubridate::ceiling_date(year, unit = "year"))

pertussis_data_tidy
```

## Removing years for which there is no data
```{r, tidy_data}
no_data_years <- c("1981":"1990", "2015":"2018") %>% 
  lubridate::ymd(truncated = 2L) 

## using an index vector to filter() the years
pertussis_data_tidy <- pertussis_data_tidy %>%
  dplyr::filter(!year %in% no_data_years)

pertussis_data_tidy$year %>%
  as.character() %>%
  as_factor() %>% levels()
```

## Join pertussis data with population data
Here we join the pertussis data with the `gapminder` data
```{r}
data("gapminder", package = "dslabs")
gapminder <- gapminder %>% as_tibble()
gapminder <- gapminder %>% 
  mutate(year =  lubridate::ymd(year, truncated = 2L)) %>%
  mutate(year = lubridate::ceiling_date(year, unit = "year"))


join <- left_join(pertussis_data_tidy, gapminder) 
```

# `{dplyr}`

## The `{dplyr}` verbes



