---
title: "Intro to R"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Data Analysis Using R - `r Sys.Date()` </div>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)

vector <- c(1:10, NA, NA, NA,  3, "marc")
vector
missing <- is.na(vector)

typeof(missing)

missing_num <- as.numeric(missing)
missing_num

sum(missing_num)
sum(missing)

sum(is.na(vector))
vector

sum(vector)

```

## R is case sensitive

*_Important note: R is case-sensitive, meaning that it will matter whether you use capital or lower case, keep this in mind if you get a cryptic warning or error!_*

## Help getting started with R

Writing R code can be challenging and espacially in the beginning you will get mind-boggling errors. R is extremely picky about capitals and every opening `(`, `[` or `{` must be matched by a closing `)`, `]` or `}`. 

Mind that `.` is the decimal placeholder!  

If you're stuck, try the help. You can get help about any R function by running `?function_name` and `?package_name` or `??package_name`, skip down to the examples and look for code that matches what you're trying to do.

Another great tool is Google: trying googling the error message, as it's likely someone else has had the same problem, and has gotten help online.

## Statistics and R

 * This course does not cover statistical applications of R. 
 * Allbeit that we will look at an example of a linear regression later on. 
 * Nowadays R is a versatile language and can be used for various applications, but it was originally designed as a language for performing statistical analyses. 
 * This is why very many statistical applcations are available for R. 
 * To start learning about statisics and R, I can highly recommend the book "Discovering Statistics Using R" by Dr. Andy Field: https://uk.sagepub.com/en-gb/eur/discovering-statistics-using-r/book236067%20

## For more on learning R in the context of statistics: start e.g. with:

http://www.statsteachr.org/ or

https://www.coursera.org/learn/statistical-inference/home/welcome or

https://www.youtube.com/watch?v=ACWuV16tdhY&index=21&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU and

https://www.youtube.com/watch?v=kvmSAXhX9Hs&index=29&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU 

and the rest of Mike Marin's lectures, that are also a really great way to start learning R.

## Packages

## List of available CRAN packages 
http://cran.r-project.org/web/packages/available_packages_by_date.html

For the course we use `tidyverse` a lot
```{r}
# install.packages("tidyverse")
library(tidyverse)
```
The `tidyverse` (in itself an R-package) is actually a collection of packages!
See: http://tidyr.tidyverse.org/ for more info

## Biological Packages
For Biological application go to http://www.bioconductor.org

Installing Bioconductor packages is different from CRAN packages
example: `{affy}` 

Each Bioconductor package has a landing page:

https://www.bioconductor.org/packages/release/bioc/html/affy.html

```{r, eval=FALSE}
## installing `{affy}` package
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite("affy")

## loading affy package and the vignettes pages of this package
# library(affy)
# browseVignettes("affy")
```

## Vignettes
Vignettes are long explanations and demos of a package. Commonly, a vignette contains examples and a workflow that shows how the package can be used and which (research) questions can be adressed with the functions and datasets in the package. It usually is a good place to start for examples. It also shows the so-called dependence of a package: it explains which other packages you would need and how the data should look to be able to work with the package the vignette belongs to. 

## Getting Help for R functions and packages
```{r, eval=FALSE}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
??dplyr
??ggplot2
?mean
??mean  # goes to the page with functions related with '.mean.'
apropos("mean") # search on more options of or alternatives for a certain function
```

## Examples and demos on functions and packages
```{r, eval=FALSE}
example(mean) # to see a worked example
demo(graphics) # demonstration of R functions
```

## Functions; naming arguments
```{r, error=FALSE, message=FALSE, warning=FALSE}
# ?rnorm
hist(rnorm(100))

seed <- c(1:4)

set.seed(seed)
q <- rnorm(n = 10000, mean = 20, sd = 2)
hist(q)
## no naming
# set.seed(1234)
qq <- rnorm(10000, 20, 2)
all(q == qq)
```

__**Write code for humans, be explicit!**__

# Data objects

## Vectors
R is an object oriented language: meaning you can create and work with (manipulate/index/access) objects. Vectors are R's elementary objects and 
come in different flavours:

 - Nummeric vector: contains only numbers: decimal separator in R is "." (decimal point) and not "," (decimal comma) as is common in the English language. 
 - Character vector: contains only "words", but words can also be numbers: "23" or other items "100%" or "$2,000.00" 
 - An integer vector: an nummeric series: 1, 2, 3 is an integer of length 3.
 - Logical: logical vectors contain only two values: "TRUE" and/or "FALSE"
 - Mixed: Vectors do not need to be of one type. They can be mixed. They can only be of one class, so this operation will induce **_coercion_**.

## Numeric vectors
```{r}
c(2,8,5) 	# combines its arguments to form a vector
nv_1 <- c(2,8,5) # assignment statement
nv_1 # view the content of the object "nv_1"
```

## Class tells you about the classification
```{r}
nv_2 <- c(8.4,5.6,10.1,13.1,2.5,7.8,15.2,3.8,20.9)
nv_2
class(nv_2)
```

# Number notations and rounding

## Scientific notations 
```{r}
big_numbers <- rnorm(10, mean = 10000000, sd = 2)
big_numbers %>% formatC(format = "e", digits = 2)

number <- c(1000)
formatC(number, format = "e", digits = 3)

large_number <- c(2.55858868688584848)


round(large_number, digits = 3)


sqrt(large_number * 1000 /200 * 2^6)  


```

## Rounding numbers
```{r}
small_numbers <- runif(10, min = 0.001, max = 0.1) %>% print()
small_numbers %>% round(digits = 2)
```

## Character vectors
```{r}
cv <- c("this is", "an", "example of", "1", "character", "vector", "with length:", "length(cv)")
cv
sum(cv)

length(cv)
class(cv)
```

## Integers
```{r}
int <- as.integer(1:5)
int
length(int)
class(int)

example_int <- c(1.4, 1.6, 7.9)
typeof(example_int)

as.integer(example_int)


```

## Logical vectors
```{r}
lv <- c(TRUE, FALSE, TRUE, TRUE)
lv
class(lv)
## logical vectors can also be converted to numeric vectors
nlv <- as.numeric(lv)
nlv
## note that coercion of a logical to a numeric vector changes the "TRUE" value to 1 and the "FALSE" value to 0
```

## Manipulating vectors
You can add/subtract/devide or use other arithmetic functions on numeric vectors
```{r} 
a <- c(1,3,5,7,9)
b <- c(2,4,6,8)

z1 <- a - b
z1


z2 <- b - a
z2
z3 <- a / b
z3
```

## Apply functions to a vector 
```{r}
z4 <- sum(a)
z4
z5 <- max(a) - max(b)
z5
```   

## Series in base-R
```{r}
numbers_times <- rep(1:4, times = 4) %>% print()
numbers_each <- rep(1:4, each = 4) %>% print()
```

## Series in `{tidyverse}`
```{r}
library(modelr)
x <- rcauchy(100)
hist(x)
```

## More `{tidyverse}` series
```{r}
seq_range(x, n = 10)
seq_range(x, n = 10, trim = 0.1)
seq_range(x, by = 1, trim = 0.1) 
```

## Make pretty `{tidyverse}` sequences
```{r}
y <- runif(100)
seq_range(y, n = 10)
seq_range(y, n = 10, pretty = TRUE)
seq_range(y, n = 10, expand = 0.5, pretty = TRUE)
```

## Indexing with [...]
You can get the individual items of a vector by using the index `[]`
```{r}
x <- c(8, 5, 10, 13, 2, 7, 15, 3, 20, 8)
x # create vector with 10 variables
length(x)
mode(x) # information on data mode (numeric, character, logic)

```

## Using the index
```{r}
x
x[3] ## creating a subset by indexing:
x[c(3, 4, 7)] # apply a simple function

mean(x[c(2,5)]) # example of a function
```

## Using the vector index `[]` some more
```{r}
i <- 5
x[c(i, i+2)] # (i = 5 & i = 7), i-th element
x[-2] # all but the ith (second) element
x[3:5] # element 3 to 5
x[x > 9] # all greater than some value
```

## Plotting and filtering elements from a vector
```{r}
set.seed(1234)
h <- rnorm(n = 1000, mean =  0, sd = 2)
plot(h)
h
max(h)
min(h)


hist(h[larger_than_0 <- h > 0 | h < -5], main = "title", xlab = "units",
     ylab = "units")





```

## Creating a logical index and apply it to a plot
```{r}
ind <- h < 0
h[ind] %>% plot()
```

## Histogram
```{r}
hist(h, breaks = 20)
```

## Truncate the distribution at -4 and 4
```{r, fig.align='center', fig.height=4, fig.width=8}
index <- h > -4 & h < 4
par(mfrow = c(1,2))
hist(h, breaks = 20)
hist(h[index], breaks = 20)
```

## Concatenate or `c()` command
The `c()` function can be used to __concatenate__, __combine__ or __create__ objects in R 
```{r}
pp <- c(x, 0, x)
pp 
sum(c(pp, c(1:5), c(1,2, c(1:10)))) # sum of elements in the 'concatenated' vector
```

## Plotting models
```{r} 
c <- c(1:7,9,11) # dependent variable (response)
d <- c(1:9) # independent variable (terms)
```

## Performing a linear correlation (response ~ terms)
```{r}
m1 <- lm(c ~ d)
m1$coefficients
```

## Model summary
```{r}
summary(m1)
```

## Make a plot 
```{r}
plot(d, c, ylim=c(0,13))
```

## Plotting the model
```{r}
plot(d, c, ylim=c(0,13), abline(m1), 
     main = "Regression Example",
     xlab = "Terms", ylab = "Response")  #abline plots the correlation model in the graph
```

## ggplot2 version
During the course we will mostly use the `ggplot2` plotting system!
ggplot2 works with dataframes, so we first need to create one
```{r} 
tbl_df <- tibble(x = d, y = c)
```

## ggplot code
```{r, results='hold'}
ggplot(data = tbl_df, aes(x = x, y = y)) +
  geom_point(colour = "darkred", size = 3) +
  geom_smooth(method = "glm", se = TRUE, formula = y ~ x, colour =         "darkblue") +
  ggtitle("Regression Example") + xlab("Terms") + ylab("Response")
```

## Combine the two in one slide
```{r, echo=FALSE, results='hold'}
source(file.path(root, "code", "panel_ggplot2_base_combined.R"))
```

## Handeling character vectors
What happens if you use arithmetic functions on character vectors and numeric vectors?
```{r, eval=FALSE}
w <- c("1", "2")
u <- c("a", "b", "c")
z6 <- w-y  
z7 <- u+w
``` 
This does not work because the vectors are of different type. We can use the paste command for this
```{r}
w <- c("1", "2")
u <- c("a", "b", "c")
z8 <- paste(u,w, sep = "_")
z8 ## because w is shorter than u, it get's recycled over u
```

## Let's clean up the workspace
```{r, eval = FALSE }
rm(list=ls())
root <- find_root_file(criterion = is_rstudio_project)
## Note: never use this in code that is meant for others!!!
```

*_The above is an effective way to clear all the items in the Global Environment, but is is not very friendly to use this in code you share with others: can you think of why?_*

## Data Structures: Lists

When using R for data analysis you will most likely work with data in a matrix, an array, a list or even more likely: a dataframe.

A matrix is a table with only numeric values. An array consists of multiple matices. A list is collection of R objects of different data type. A dataframe is a table with variable names in the first row and observations in the consecutive rows. The columns in a dataframe represnet different variables.  

The dataframe and the list are the most widely used datastructures when considering experimental Biological data. 

## Create a List
```{r}
lst <-list(first_names = c(male = "Fred", female = "Mary"), no.children = 3, child_ages=c(4,7,9),
child_names = c("Suzy", "Marvin", "Jane"), address = c("Pandmonium Alley 114, Chaosville"),
marital_status = TRUE)
length(lst) # number of elements in the list
head(lst, 2)
```

## Accessing items in a list
```{r}
lst[[1]] # 1st element of List
lst[[3]][2] # second item of third element
names(lst) # named elements in this list
lst$child_names # pull "named" elements from a list using `$` operator
```

## `str()` gives you the structure of an object
```{r}
str(lst) # display structure of lst
``` 
The new RStudio Interface also enables interactive exploration of R-objects (demo)

## Selecting single elements in a list
To select a single element from a variable in a list
```{r}
lst$child_ages[3] 
lst[[6]][2] # returns the value of the second element for your variable
``` 

## Looping over lists
```{r}
purrr::map(lst, is.na)
```
## The Dataframe
The dataframe is the most widely used data structure in the context of experimental biology and chemistry. **Remember "Tidy data!"**   

## Creating a data frame from individual vectors
```{r}
people_df <- data.frame(age=c(24, 27, 19, 34),      
                       sex=c("F","F","M", "M"), 
                       weight=c(64, 55, 80, 70),
                       names = c("Christa", "Suzan", 
                                 "Matt", "John"))
head(people_df)
``` 

## Or the 'tidyverse' way
```{r}
people_tbl <- tibble::tibble(age=c(24, 27, 19, 34),      
                       sex=c("F","F","M", "M"), 
                       weight=c(64, 55, 80, 70),
                       names = c("Christa", "Suzan", 
                                 "Matt", "John"))
people_tbl
```

## Viewing the contents of a dataframe
```{r}
summary(people_df)
table(people_df)

head(people_df) 			# gives the content of the data frame
names(people_df) 
str(people_df)

people_df$age # gives the content of the variable "age" from the data frame ""

``` 

## Using index on dataframes
Using the index "[]" on a dataframe is a bit tricky. The dataframe always consists of rows and columns. Indexing a dataframe goes like:

`dataframe[row number(s), column number(s)]`

```{r}
people_df$age[1] 	# first element of this vector
people_df[,2] 	# content of 2nd variable (column) which is a character vector -> factor
people_df[1,] 	# content of the 1st row
# multiple indices
people_df[2:3, c(1,3)] # remember to use c

```

## Import data into R
```{r}
library(tidyverse)
path_to_gender_age_data <- file.path(root, "data", "gender.txt")
gender_age <- read_delim(path_to_gender_age_data,
                         delim = "/")
```

## Inspecting the dataframe
```{r}
# getting the first few rows 
head(gender_age)
# getting information on the variable, the dimensions
str(gender_age)
# selecting a variable
gender_age$gender
gender_age$age
gender_age[1:6,2]
``` 

## read_csv
CSV is a format of a data file that uses commas or semicolons as separators for the columns.

```{r}
library(readr)
skin <- read_csv(paste0(root, "/data/skincolumns.csv")) 
str(skin)
```

## Inspecting the skin dataframe
```{r}
head(skin)	 # content of the data frame
dim(skin)
attributes(skin)
summary(skin)
## ?read_csv 	 # help on the function
```

## `skin` dataset contains an NA, some functions do not work with NAs:
```{r}
mean(skin$`Genotype A`)
mean(skin$`Genotype B`)

# to remove the NA
skin_noNA <- na.omit(skin)
mean(skin_noNA$`Genotype B`)
``` 

## Smoking example with multiple variables
Create data frame with 4 variables:
gender, 
smoke, 
age,
weigth

## Create individual vectors
```{r}
set.seed(6000) # set it at the same number, then we will all produce the same output
gender<-c(sample(c(1,2), size=1000, replace=TRUE))
smoke<-c(sample(c(1,2), size=1000, replace=TRUE))
age<-c(sample(c(1,2,3), size=1000, replace=TRUE))
weight<-round(rnorm(1000, mean=65, sd=10),1)
smoking <- data.frame(cbind(gender, smoke, age, weight))
```

## `head(x, ...)`
```{r}
head(smoking, 5)
smoking[1:5,]
``` 

## Replace numbers by descriptive labels 
```{r}
# replace numbers by characters
smoking$genderf=factor(smoking$gender, labels=c("female","male"))
table(smoking$genderf,smoking$gender)
smoking$smokef=factor(smoking$smoke, labels=c("Y","N"))
table(smoking$smokef,smoking$smoke)
smoking$agecf=factor(smoking$age, labels=c("A: <15","B: 15-30","C: >30"))
table(smoking$agecf,smoking$age)
summary(smoking)

``` 

## Subsetting data frame; the easier {dplyr} way
```{r}
library(tidyverse)
names(smoking)
smoking_male <- smoking %>% filter(genderf == "male")
smoking_age_gender <- smoking %>% select(genderf, agecf)

```

## Or the base-R way
```{r}
library(tidyverse)
names(smoking)
# create a logical index
smoking_male_index <- smoking[, "genderf"] == "male" 
smoking_male <- smoking[smoking_male_index, ]
# slecting only age and gender
smoking_age_gender <- smoking_male[, c(1,3)]
```

## Summarizing data
For summarizing data there are many ways in R (as is the case with many operations). I find the dplyr way the most intuitive. Using dplyr has the advantage that you can easily built on existing function and examples, because the code is faily easy to understand.

## The pipe or %>% operator

 * One thing that makes dplyr great is that it works with the `%>%` symbol. Which is in programming jargon also called the "pipe" symbol.
 * The pipe symbol takes the previous result of an operation and put it in the next.
 * The pipe symbol can replace the use of annoying round brackets: or "(parentheses)" 

## Example 
If we want to calculate the sum, of the square root of the mean, of two nummeric vectors (each of length = 1000) we could do:
```{r}
library(dplyr)
set.seed(12345)    
zzz <- sqrt(mean(x <- rnorm(10000, mean = 34, sd = 2))) +
sqrt(mean(y <- rnorm(10000, mean = 23, sd = 4)))
zzz
```

The above is almost impossible to read because of all the round brackets, note the closing brackets: 3 of them!!

## let's try the dplyr way with the %>% (pipe) operator
```{r}
set.seed(12345)
qqq <- rnorm(10000, mean = 34, sd = 2) %>% 
  mean() %>% 
  sqrt() + rnorm(10000, mean = 23, sd = 4) %>% 
  mean() %>% 
  sqrt()
qqq

zzz == qqq
```

The dplyr way is a lot better is it not? Do you get what the above (dplyr-way) lines do?

## The dplyr way explained 
 1. the result of the fist vector (1000 random numbers, with mean of 34 and a sd of 2 is piped into the function mean, which calculates the mean of these 1000 numbers (which will be around 34)
 2. From that the square root is taken by piping the mean in the function sqrt()
 3. The result of the firt vector is added to the calculation of the second vector (which is done the same way as the first).
 4. The result is qqq
 5. Is qqq equal to zzz, which we calculated the "old-fashioned" mathematical way? YES! 

## Let's apply the above to our smoking dataset

 * Usually, if we want to make a graph we need some sort of summarizing variable that indicates wat is going on, e.g. a group mean and standard deviation.

 * From the smoking dataset we will genrate a new dataframe with average weight, grouped by gender, age and grouped by smoking or non smoking. We will use dplyr verse to do so.

 * I will also show you how to sort the data, let's say to decreasing weight, by using the dplyr function 'arrange'. 

## Summary of the smoking data
```{r}
## it is handy to have the names of the dataset variables at hand when writing pipes

names(smoking)
str(smoking)
```

## Example `pipe`
```{r}
smoking_weight <- smoking %>% select(genderf, smokef, weight, agecf) %>%
  group_by(genderf, smokef, agecf) %>% summarise(mean_weight = mean(weight)) %>%
  arrange(desc(mean_weight))

head(smoking_weight)
```

## Looking in detail at the pipe

Let's look at what is going on above.

The smoking_weight dataset contains newly created, more descriptive denominators, we will use those. 
 
 1. The first step is to inspect the dataframe to see if the grouping variables are set to "factor" 
 2. The second step is to select the proper variables to work with
 3. Than the data is grouped by the genderf, agecf and by the smokef variables
 4. Summarise creates a new variable `mean_weight` that is what it says: the mean weight for the groups defined.
 5. Finally, the data is sorted, according the newly created mean_weight variable, in decending order "(dec(mean_weight))". If we want ascending order we can just use arrange(mean_weight). Arrange has the default to order the variable in ascending order.