---
title: "Hands-on Part 1 - Introduction to R and RStudio"
subtitle: "SURF Research Bootcamp"
author: "Marc A.T. Teunis, Jan-Willem Lankhaar, Jonathan de Bruin"
date: "November 2nd, 2018"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width = 5, 
                      fig.height = 3)

```

```{r, define_root, include=FALSE}

## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)

```

```{r, include=FALSE}

library(tidyverse)

```

## Getting to know your neighbour

Spend 4 minutes explaining to your neighbour:

 - What brings you here? What are your expectations?
 - What characterizes you regarding R: `Beginner`, `Intermediate`, `Advanced`, `Superman`/`Superwoman`?
 - What type of data do you normally work with? 
 - What type of problems are you solving or planning to solve with R?

## **Wetting your appetite...**

If you're a(n) (absolute) beginner: 

Take a deep breath..., 

keep calm..., 

and carry on...

## Creating a data table (`tibble`)
```{r}

data_example <- tibble(numbers = c(1:9, -999), 
                 more_numbers = c(1:5, -999, -999, 8:10),
                 some_letters = LETTERS[1:10])
data_example

```

## Tibbles

"Tibbles are data frames, but they tweak some older behaviours to make life a little easier. R is an old language, and some things that were useful 10 or 20 years ago now get in your way. It’s difficult to change base R without breaking existing code, so most innovation occurs in packages."

From the book: "R for Data Science" by Garrett Grolemund & Hadley Wickham

## **Discuss with your neighbour** 

**What is 'wrong' with the following R code?^1^**

 * This code is supposed to replace the number `-999` for a missing value (`NA`)
 * Find at least 3 'mistakes'
 * What would you do to prevent this kind of mistakes?

```{r, eval=FALSE}

data_example$numbers[data_example$numbers == 999] <- NA
data_example$numbers[data_example$more_numbers == -999] <- NA
data_example$numbers[data_example$some_letters == 999] <- NA
data_example$more_numbers[data_example$more_numbers == 998] <- NA

```

^1^Example (adapted from): Hadley Wickham, *Expressing yourself using R* Lecture  - <https://richmedia.lse.ac.uk/methodologyinstitute/20170814_ExpressYourselfUsingR.mp4>

## Programming with functions

When working with R you can:

Solve the same problem over and over again (and waste a lot of time in finding the errors)
<br>
<center>**OR**</center>
<br>
Write small pieces of code (functions) that are generalizable for re-use in the future

## Let's write a function that 

**replaces a designated number or string (na_string) in a vector (x) with a missing value (`NA`)**

```{r}

replace_na_string_with_na <- function(x, na_string){
  
  x[x == na_string] <- NA
  return(x)

}

```

## Apply the function to a single column in our `data_example` tibble
```{r}

data_example$numbers
replace_na_string_with_na(data_example$numbers, na_string = "-999")

```

## Take home message

 Instead of copying pieces of code over and over again...
 
 ...write re-usable pieces of code (functions)

We hope this small example inspired you to write (more of) your own functions

# Let's create some R objects
### Switch to RStudio RMarkdown document "part1_intro_to_rstudio_and_r.Rmd"

## _'numeric'_ vectors
```{r}

numbers <- c(1.3, 3.6, 7.8)
numbers
sum(numbers)

integer_vector <- c(1:5)
integer_vector

```

## What happens to a 'mixed' vector?
```{r, mixed}

vector <- c(1:3, NA, NA, NA,  3, "Python is great!", "R is greater!")
vector
is.na(vector) 

```

## Type of object
```{r}

class(numbers)
class(vector)
class(integer_vector)

```

## Getting number of `NA`s
```{r}

is.na(vector) ## how many TRUEs
sum(is.na(vector)) ##  how many NAs

```

## Mathematical operations on a `character` vector
```{r, eval=FALSE}

sum(vector) ## Warning sum on character

```

## Let's create some data 
```{r}

normals <- rnorm(n = 100, mean = 5, sd = 1.3)
mean(normals)
## change number of observations to 10000 and rerun

```

## Plot histogram
```{r}

normals_tbl <- as_tibble(normals) 

ggplot(data = normals_tbl, aes(x = value)) +
  geom_histogram(bins = 50)
## change the `bins` argument to see what happens
## change the n = ... argument in the code above and rerun 

```

## **Discuss with your neighbour**

 * How would you turn the above plot code into a function that takes two arguments: `bins` and `df`?
 * Try it out using the following skeleton:
```{r, eval=FALSE}

plot_histogram <- function(df, bins){
 ## Write your code here, use the example above
 fig <- df %>% ggplot(....)
 ....
 ....
 ....

 return(fig)
}
## test function
plot_histogram(df = normals_tbl, bins = 40)

```

## R is case-sensitive!

**Important note**

_R is case-sensitive, i.e. it will regard names with different cases as different names._

```{r, eval=FALSE}

x <- 1
print(x)
print(X)
X <- "this is a different X"
print(x)
print(X)

```

_Keep this in mind if you get a warning or error on `'object '...' not found!`_

## Programming (in R) can be frustrating

 >- Writing (R) code can be challenging 
 >- You will get mind-boggling errors
 >- R is extremely picky about capitals
 >- Every opening `(`, `[` or `{` must be matched by a closing `)`, `]` or `}` 

## Decimals

Mind that in R `.` is the decimal separator!

```
# Example
pi <- 3.14159

# NOT pi: 3,14159
```

## Getting help

 >- If you're stuck, try the help (`?` or `??` or `help()`) function. 
 >- You can get help on any R function by running `?function_name` and `?package_name` 
 >- or `??package_name`, skip down to the examples and look for code that matches what you're trying to do.
 >- Other great tools are `Google` and `Stack Overflow`: google an error message. It's very likely that someone else encountered the same problem and published a solution.

## Getting help examples
```{r, eval=FALSE}

?dplyr
?ggplot2
?mean

```

## For more on learning R, start e.g. with:

https://r4ds.had.co.nz/ !!

http://www.statsteachr.org/ or

https://www.coursera.org/learn/statistical-inference/home/welcome or

https://www.youtube.com/watch?v=ACWuV16tdhY&index=21&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU and

https://www.youtube.com/watch?v=kvmSAXhX9Hs&index=29&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU 

and the rest of Mike Marin's lectures, that are also a really great ways to start learning R.

Or try datacamp.com

# Packages

## List of available CRAN packages 
http://cran.r-project.org/web/packages/available_packages_by_date.html

For this workshop we mainly use the `{tidyverse}` 
```{r, echo=FALSE}

knitr::include_graphics(path = file.path(root,
                                         "images",
                                         "tidyverse_sticker.png"))

```
```{r}

# install.packages("stringr")
library(tidyverse)

```

## `{tidyverse}`
See: http://tidyverse.tidyverse.org/ for more info
The `{tidyverse}` package includes (among others):

 >- Wrangling: `{dplyr}`, `{tidyr}` 
 >- Reading data: `{readr}`, `{readxl}`
 >- Functional prgramming: `{purrr}`
 >- Models: `{broom}`
 >- Strings and regular expressions: `{stringr}`
 >- Factors: `{forcats}`
 >- Time (series): `{lubridate}`

For a full list, type `tidyverse_packages()` in the console, try it now!


## Function calls: to name or not to name

When you use a function (i.e. you 'call the function') you can either explicitly specify which value you want to use for which argument:
```{r, error=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
q <- rnorm(n = 10000, mean = 20, sd = 2)
```
... or you can do so implicitly (i.e. by position):
```{r, eval=FALSE}
qq <- rnorm(10000, 20, 2)
```

__**Write code for humans AND computers, be explicit!**__

# R (Data) objects

## Concatenate or `c()` command
The `c()` function can be used to __concatenate__, __combine__ or __create__ objects in R 
```{r}
x <- c(1, 2, 3, 4, 5)
pp <- c(x, 0, x)
length(pp) 
sum(pp) # sum of elements in the 'concatenated' vector (2 x 15)
```

## Data Structures: Lists {.build}

When using R for data analysis you will most likely work with data in a matrix, an array, a list or even more likely: a dataframe.

 >- A matrix is a table with only one value type (usually numbers) 
 >- An array consists of multiple matices
 >- A list is collection of R objects of different data type and can be of different length 
 >- A dataframe is a table with variable names in the first row and observations in the consecutive rows. The columns in a dataframe represent different variables. All variables have equal lenght  

_The dataframe and the list are the most widely used data structures for experimental data._ 

## Create a List
```{r}

lst <-list(first_names = c(male = "Fred", 
                           female = "Mary"), 
           no.children = 3, 
           child_ages=c(4,7,16),
           child_names = c("Suzy", "Marvin", "Jane"), 
           address = c("Pandamonium Alley 114, Chaosville"),
           marital_status = TRUE)

length(lst) # number of elements in the list
names(lst)

```


## Printing the contents of the list
```{r}

lst # we can explore the contents interactively in RStudio interface

```

## Accessing items in a list
```{r}

lst[[1]] # 1st element of List
lst[[3]][2] # second item of third element
names(lst) # named elements in this list

```

## Using `$` on named elements in a list
```{r}

lst$child_names 

```

## `glimpse()` gives you the structure of an object
```{r}

glimpse(lst) # display structure of lst

``` 
The new RStudio Interface also enables interactive exploration of R-objects (demo)

## Selecting single elements in a list
To select a single element from a variable in a list
```{r}

lst$child_ages[3] 
lst[[5]][1] # returns the value of the first item of the 5th element 

``` 

## The data frame
The data frame is the most widely used data structure 

## Creating a data frame from individual vectors
```{r}

people_df <- data.frame(age = c(24, 27, 19, 34),      
                        sex = c("F","F","M", "M"), 
                        weight = c(64, 55, 80, 70),
                        names = c("Christa", "Suzan", 
                                  "Matt", "John"))
head(people_df)

``` 

## Or the 'tidyverse' way
```{r}

people_tbl <- tibble(age = c(24, 27, 19, 34),      
                     sex = c("F","F","M", "M"), 
                     weight = c(64, 55, 80, 70),
                     names = c("Christa", "Suzan", 
                               "Matt", "John"))
people_tbl

```

## Viewing the contents of a data frame
```{r, eval=FALSE}

summary(people_df)
table(people_df)

head(people_df) 			# gives the content of the data frame
names(people_df) 
str(people_df)

people_df$age # gives the content of the variable "age" from the data frame

``` 

## Using index on dataframes
Using the index "[]" on a dataframe is a bit tricky. The dataframe always consists of rows and columns. Indexing a dataframe goes like:

`dataframe[row number(s), column number(s)]`

```{r, eval=FALSE}

people_df$age[1] 	# first element of this vector
people_df[ , 2] 	  # content of 2nd variable (column) which is a character                      vector -> factor
people_df[1 , ] 	  # content of the 1st row
                  # multiple indices
people_df[2:3, c(1,3)] # remember to use c()

```

## Import data into R
```{r}

library(tidyverse)
path_to_gender_age_data <- file.path(root, "data", "gender.txt")
gender_age <- read_delim(path_to_gender_age_data,
                         delim = "/")

```

## read_csv
CSV is a format of a data file that uses commas or semicolons as separators for the columns.

```{r}

library(readr)
file_location <- file.path(root, "data", "skincolumns.csv" )
skin <- read_csv(file_location) 
head(skin, 3)

```

## `skin` dataset contains an NA, some functions do not work with NAs:
```{r}
mean(skin$`Genotype A`)
mean(skin$`Genotype B`)
mean(skin$`Genotype B`, na.rm = TRUE)
``` 

# Optional

Or move on to part 2

## Manipulating vectors
You can add/subtract/devide or use other arithmetic functions on numeric vectors
```{r} 
a <- c(1,2,4,6)
b <- c(2,4,6,8)

a + b
b - a
a / b
```

## Apply functions to a vector 
```{r}
sum(a)
max(a) - max(b)
```   

## Indexing with [...]
You can get the individual items of a vector by using the index `[]`
```{r}
x <- c(8, 5, 10, 13, 2, 7, 15, 3, 20, 8)
x # create vector with 10 variables
length(x)
```

## Using the index
```{r}
x[3] ## creating a subset by indexing:
x[c(3, 4, 7)] # apply a simple function
mean(x[c(2,5)]) # example of a function
```

## Demo brush "`Clean the  Global Environment`"

