---
title: "Hands-on - Visualizations with R"
subtitle: "SURFsara Research Bootcamp"
author: "Marc A.T. Teunis; Jan-Willem Lankhaar; Jonathan de Bruin"
date: "November 6th 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE, 
                      fig.width = 6, fig.height = 4)

## ggplot theme set
library(ggplot2)
theme_set(theme_bw())
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Get materials
The materials in this document are hosted on Github:

https://github.com/uashogeschoolutrecht/surfR

To get them locally (in RStudio)

1. Start a new project (`File` --> `New Project`)
2. Choose `Version Control` --> `Git`
3. Paste the above url in the `Repository URL` field
4. Click `OK`
5. The materials will be cloned to your account or local computer

The rendered HTML document is also hosted on our RStudio::CONNECT server at:

http://connect3.hudatascience.nl:81/surfr_viz/part2_demo_visualizations.html

## Reference
https://simplystatistics.org/2018/01/22/the-dslabs-package-provides-datasets-for-teaching-data-science/

https://github.com/rbind/simplystats

[link to free online book "r4ds":](https://r4ds.had.co.nz/)
"R for Data Science", Hadley Wickham & Garret Grolemund


## Gapminder
https://www.gapminder.org/tools/#$state$time$value=2018;;&chart-type=bubbles 

## ggplot2
The ggplot2 package in R is the best plotting system for R. It's syntax is an implementation of the 'grammar of graphics' and all plots in this demo are created by using ggplot2.

```{r, echo=FALSE}
knitr::include_graphics(path = file.path(root, "images",
                                         "hex-ggplot2.png"))
```

## Citations
```{r}
citation(package = "ggplot2")
citation(package = "tidyverse")
citation(package = "dslabs")
```

## Installing data package 
```{r, eval=FALSE}
install.packages("dslabs")
```

## Loading the data package `{dslabs}` and other packages used
```{r}
library(tidyverse)
library(dslabs)
library(lubridate)
```


## Datasets 
Included in the `{dslabs}` package
```{r}
data(package="dslabs") %>% print()
```

## Getting original data wrangling scripts 
Included also in the `{dslabs}` package
```{r}
list.files(system.file("script", package = "dslabs"))
```

## `{ggplot2}` is the core visualization package
```{r, echo=FALSE}

knitr::include_graphics(path = file.path(root, "images",
                                         "hex-ggplot2.png"))
```

## Demo dataset 
```{r}
data("gapminder", package = "dslabs")
## ?gapminder for more info on the variables in the dataset
```

The gapminder dataset contains a number of measurements on health and income outcomes for 184 countries from 1960 to 2016. It also includes two character vectors, OECD and OPEC, with the names of OECD and OPEC countries from 2016. 

## Inspecting the `gapminder` dataset with R
```{r}
gapminder <- gapminder %>% as_tibble()
gapminder %>% head(2)
names(gapminder)
```

## A very simple example to start with
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point()
```

## This is a very dense plot
We call this 'overplotting'.

This can be fixed in several ways:

 >- Reducing the transparency of data points  
 >- Mapping colour to a variable (continuous or categorical)
 >- Reduce the data in the plot
 >- Mapping a shape to a variable
 >- Add noise (`"jitter"`) to points
 >- Facetting - create panels for 'categorical' or so-called 'factor' variables in R
 >- Summarize the data
 >- Displaying a model / relationship that represents the data (and not sho the actual data itself) 
 >- Or any combination of the above strategies

__Basically you map an `aesthetic` (`aes()`) to a variable__
 
Let's go over these overplotting methods one by one 

## 1. Overplotting: Reducing transparency (`alpha`) of points or lines in the data
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(alpha = 0.1)
```

## 2. Mapping colour to a variable (continuous or categorical)
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent))
```

## or combined with `alpha`
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.1) +
  guides(colour = guide_legend(override.aes = list(alpha = 1)))

```

## **Do it yourself:**

 - Try adjusting some of the arguments in the previous `ggplot2` call. 
For example, adjust the `alpha = ...` or change the variable in `x = ...`, `y = ...` or `colour = ...`
 - `names(gapminder)` gives you the variable names that you can change
 - Show and discuss the resulting plot with your neighbour
 - What do you think this part does: 
 
 `guides(colour = guide_legend(override.aes = list(alpha = 1)))`
 
 - Try to find out by disabling with `#`

## 3. Reduce the data in the plot
```{r}
reduce_data_plot <- gapminder %>% 
  dplyr::filter(continent == "Africa" | continent == "Europe") %>%
  
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.2) +
  ## override the alpha setting for the points in the legend:
  guides(colour = guide_legend(override.aes = list(alpha = 1))) 


```

## Plot
```{r}
reduce_data_plot
```

## **Discuss with you neighbour:**

 - What does the the `aes()` part of the `geom_point()` do?
 - Compare the code below with the code above, can you spot the difference, what is the advantage of the code below?
 
```{r, eval=FALSE}
reduce_data_plot <- gapminder %>% 
  filter(continent == "Africa" | continent == "Europe") %>%
  
  ggplot(aes(x = fertility,
             y = life_expectancy, colour = continent)) +
  geom_point(alpha = 0.2) +
  ## override the alpha setting for the points in the legend:
  guides(colour = guide_legend(override.aes = list(alpha = 1))) 
```

## 4. Mapping a shape to a variable
```{r}
## or e.g. show only two years and map a shape to continent
shape_plot <- gapminder %>% 
  dplyr::filter(continent == "Africa" | continent == "Europe",
         year == "1960" | year == "2010") %>%
  
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = as_factor(as.character(year)), 
                 shape = continent), 
             alpha = 0.7)
```

## **Do it youself**

 - Try removing the `as_factor(as.character(year))` call and replace this by only `year` above and rerun the plot, what happened?

## Plot
```{r}
shape_plot
```

## 5. Facetting

Create panels for 'categorical' or so-called 'factor' variables in R
```{r}
facets_plot <- gapminder %>% 
  dplyr::filter(continent == "Africa" | continent == "Europe",
         year == "1960" | year == "2010") %>%
  
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.5) +
  facet_wrap(~ year)

```

## Plot
```{r}
facets_plot
```

## 6. Summarize the data
```{r}
library(ggrepel)

years <- c("1960", "1970", "1980", "1990", "2000", "2010")

summarize_plot <- gapminder %>% 
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_life_expectancy = mean(life_expectancy),
            mean_fertility = mean(fertility)) %>%
  ggplot(aes(x = mean_fertility,
             y = mean_life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.7) 

```

## Plot
```{r}
summarize_plot
```

## Adding labels to the points with `{ggrepel}`
```{r}
library(ggrepel)

years <- c("1960", "1970", "1980", "1990", "2000", "2010")

labels_plot <- gapminder %>% 
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_life_expectancy = mean(life_expectancy),
            mean_fertility = mean(fertility)) %>%
  ggplot(aes(x = mean_fertility,
             y = mean_life_expectancy)) +
  geom_point(aes(colour = continent), alpha = 0.7) +
  geom_label_repel(aes(label=year), size = 2.5, box.padding = .5)
  
```

## Plot
```{r}
labels_plot
```

## 7. Displaying a model / relationship that represents the data (and not show the actual data itself) 
```{r}
## Model
lm <- gapminder %>% lm(formula = life_expectancy ~ fertility)

correlation <- cor.test(x = gapminder$fertility, 
                        y = gapminder$life_expectancy, 
                        method = "pearson")

# save predictions of the model in the new data frame 
# together with variable you want to plot against
predicted_df <- data.frame(gapminder_pred = predict(lm, gapminder), 
                           fertility = gapminder$fertility)

```

## Add model to plot
```{r}
model_plot <- gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
#  geom_point(alpha = 0.03) +
  geom_line(data = predicted_df, aes(x = fertility, 
                                     y = gapminder_pred),
            colour = "darkred", size = 1)
```

## Plot
```{r}
model_plot
```

## Plotting statistics to the graph with the `{ggpubr}` package
```{r, echo=FALSE}
library(ggpubr)
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(alpha = 0.02) +
  
  geom_line(data = predicted_df, 
            aes(x = fertility, 
                y = gapminder_pred),
            colour = "darkred", size = 1) +
  stat_cor(method = "pearson", label.x = 2, label.y = 30) +
  theme_bw()

```

## Using a smoother `geom_smooth` to display potential relationships
```{r}
gapminder %>% 
  ggplot(aes(x = fertility,
             y = life_expectancy)) +
  geom_point(alpha = 0.02) +
  geom_smooth(method = "lm") +
  stat_cor(method = "pearson", label.x = 2, label.y = 30) +
  theme_bw()
```

## Recap **Discuss with your neighbour**
Which tricks can we use to reduce the dimensionality of the plotted data (prevent overpltting)?

Try listing at least 6 methods:

## Relation between `gdp`, Gross Domestic Product and `infant_mortality` rate. 
https://en.wikipedia.org/wiki/Gross_domestic_product
Wikipedia: Gross Domestic Product (GDP) is a monetary measure of the market value of all the final goods and services produced in a period of time, often annually or quarterly. Nominal GDP estimates are commonly used to determine the economic performance of a whole country or region, and to make international comparisons. 
```{r}
gdp_infant_plot <- gapminder %>%
  dplyr::filter(continent == "Europe" | continent == "Africa") %>%
  ggplot(aes(x = gdp, 
             y = infant_mortality)) +
  geom_point() 

```

## Plot
```{r}
gdp_infant_plot
```

## Adding a bit of colour
The figure above does not provide any clue on a possible difference between Europe and Africa, nor does it convey any information on trends over time.

```{r}
colour_to_continent <- gapminder %>%
  dplyr::filter(continent == "Europe" | continent == "Africa") %>%
  ggplot(aes(x = gdp, 
             y = infant_mortality)) +
  geom_point(aes(colour = continent))

```

## Plot
```{r}
colour_to_continent
```

## Adding facets
Let's investigate whether things have improved over time. We compare 1960 to 2010 by using a panel of two figures. Adding simply `facet_wrap( ~ facetting_variable)` will do the trick.

## **Discuss with your neighbour:**

Without looking ahead try to contruct a plot that conveys information on the `gdp` per continent, over time. Try to recycle some of the examples above.  































```{r}
facets_gdp_infant <- gapminder %>%
  dplyr::filter(continent == "Europe" | continent == "Africa",
                year == "1960" | year == "2010") %>%
  ggplot(aes(x = gdp, 
             y = infant_mortality)) +
  geom_point(aes(colour = continent)) +
  facet_wrap(~ year) +
  theme(axis.text.x = element_text(angle = -90, hjust = 1))
  

```

## Plot
```{r}
facets_gdp_infant
```

## Mapping to continuous variables
So far we have been mapping colours and shapes to categorical variables. You can also map to continuous variables though.
```{r}
continuous <- gapminder %>%
  dplyr::filter(country == "Netherlands" | 
                country == "China" |
                country == "India") %>%
  dplyr::filter(year %in% years) %>%
  ggplot(aes(x = year,
         y = life_expectancy)) +
  geom_point(aes(size = population, colour = country)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  geom_line(aes(group = country)) +
  theme_bw()
```

## Plot
```{r}
continuous
```

## **Discuss with your neighbour**
Try plotting the `infant_mortality` against the filtered years for the same countries as the code above (Netherlands, India, China), recycling some of the code above. Discuss the resulting graph in the light of the life_expectancy graph, what do you think about the the developments in China? 

Want to know more? see: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4331212/
Babxiarz, 2016

## **Discuss with your neighbour**
Analyze the following code chunk: try running line by line to see what happens:

 - How many observations are we plotting here?
 - How many variables are we plotting?
 - Try adding or removing variables to the `group_by()` statement, what happens if you do?

## Summarize per continent and sum population
```{r}
population_plot <- gapminder %>% 
  dplyr::group_by(continent, year) %>%
  dplyr::filter(year %in% years) %>%
  summarise(sum_population = sum(population)) %>% 
  ggplot(aes(x = year, 
             y = sum_population)) +
    geom_point(aes(colour = continent)) +
    geom_line(aes(group = continent,
                  colour = continent))
```

## Plot
```{r}
population_plot
```

## Ranking data
```{r}
ranking_plot <- gapminder %>%
  dplyr::filter(continent == "Europe",
                year == 2010) %>%
  ggplot(aes(x = reorder(as_factor(country), population),
             y = log10(population))) +
  geom_point() +
  ylab("log10(Population)") +
  xlab("Country") + 
  coord_flip() +
  geom_point(data = filter(gapminder %>%
  dplyr::filter(continent == "Europe",
                year == 2010), population >= 1e7), colour = "red")

```

## Plot
```{r, fig.height=8, fig.width=6}
ranking_plot
```

## Let's look at a time series
We filter for "Americas" and "Oceania" and look at `life_expectancy` over the years.
```{r}
## without summarizing for countries
gapminder$continent %>% as_factor() %>% levels()
gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  ggplot(aes(x = year,
             y = life_expectancy)) +
  geom_line(aes(group = continent,
                colour = continent))
```  

**Obviously something went wrong here. Please, discuss with your neighbour what you think happened or needs to be done to fix this (without looking ahead ;-) )**
























## Grouping
We can see what happened if we plot individual datapoints
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  ggplot(aes(x = year,
             y = life_expectancy)) +
  geom_point(aes(colour = country)) +
  theme(legend.position="none") +
  facet_wrap( ~ continent) +
  theme(legend.position="none") 
  
```

## Summarizing time series data
```{r}
gapminder$continent %>% as_factor() %>% levels()
gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  group_by(continent, year) %>%
  summarise(mean_life_expectancy = mean(life_expectancy)) %>%
  ggplot(aes(x = year,
             y = mean_life_expectancy)) +
  geom_line(aes(group = continent,
                colour = continent)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Statistical proof?
```{r}

df <- gapminder %>% 
  dplyr::filter(continent == "Americas" | continent == "Oceania") %>%
  group_by(continent, year)

model <- aov(data = df, life_expectancy ~ continent * year)
anova(model)

```

Some remarks on the above Two-way ANOVA:

 - Repeated measures / multilevel models might be more appropriate here (paired / nested designs)
 - We did not perform any check on assumptions
 - We performed our analysis on only part of the data
 
## One more option: categorical values and "jitter"
Sometimes you have overlapping plots and adding transparency with `alpha()` or mapping colour to underlying categorical values is not working because there are simple to many points overlapping

Let's look at an example
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  group_by(continent) %>%
  dplyr::filter(year %in% years) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  geom_point(aes(colour = country)) +
  theme(legend.position="none")
  
```

In such cases it can be helpfull to add some noise to the points (`position = "jitter"`) to reduce overlapping. This can be a powerfull approach, especially when combined with setting `alpha()`
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
    group_by(continent) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  geom_point(aes(colour = continent), position = "jitter") 
  
```
 
## Bar chart 
It would be nice to know what the mean child mortality is for both continents
```{r}
gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_infant_mortality = mean(infant_mortality, na.rm = TRUE)) %>% 
  ggplot(aes(x = year,
             y = mean_infant_mortality)) +
  geom_col(aes(fill = continent), position = "dodge") 
  
```

## Adding summary data to an existing plot
Now that we have the mean infant mortality for each year for the two continents, let's add that data to the previous dot plot where we used jitter
```{r}
mean_inf_mort <- gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
  group_by(continent, year) %>%
  summarise(mean_infant_mortality = mean(infant_mortality, na.rm = TRUE))

gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
    group_by(continent) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  geom_point(aes(colour = continent), position = "jitter") +

## summary data added to previous 
  geom_line(data = mean_inf_mort, aes(x = year, 
                                      y = mean_infant_mortality, 
                                      colour = continent),  size = 2)


```

## Filter data from a graph
In the figure above we can observe a number of countries in 'Americas' continent that have a child mortality that are above the average (over the years) of 'Africa'. Which countries are this?

```{r}
library(ggiraph)

gapminder$country <- 
  str_replace_all(string = gapminder$country, 
                pattern = "'", 
                replacement = "_")


interactive_inf_mort <- gapminder %>% 
  dplyr::filter(continent == "Americas" |
                continent == "Africa") %>%
  dplyr::filter(year %in% years) %>%
    group_by(region, country) %>%
  ggplot(aes(x = year,
             y = infant_mortality)) +
  
  geom_point_interactive(aes(tooltip = country, colour = region), position = "jitter") +
  
#  geom_point(aes(colour = continent), position = "jitter") +

## summary data added to previous 
 geom_line(data = mean_inf_mort, aes(x = year, 
                                      y = mean_infant_mortality, 
                                      colour = continent, group = continent),  size = 2
            )

interactive_inf_mort

gapminder$country %>% as_factor() %>% levels()
ggiraph(ggobj = interactive_inf_mort)

```






 
## A more complicated example (for showing the capabilities of ggplot2)
```{r}
west <- c("Western Europe","Northern Europe","Southern Europe",
          "Northern America","Australia and New Zealand")

gapminder <- gapminder %>%
  mutate(group = case_when(
    region %in% west ~ "The West",
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region %in% c("Caribbean", "Central America", "South America") ~ "Latin America",
    continent == "Africa" & region != "Northern Africa" ~ "Sub-Saharan Africa",
    TRUE ~ "Others"))
gapminder <- gapminder %>%
  mutate(group = factor(group, levels = rev(c("Others", "Latin America", "East Asia","Sub-Saharan Africa", "The West"))))

filter(gapminder, year%in%c(1962, 2013) & !is.na(group) &
         !is.na(fertility) & !is.na(life_expectancy)) %>%
  mutate(population_in_millions = population/10^6) %>%
  ggplot( aes(fertility, y=life_expectancy, col = group, size = population_in_millions)) +
  geom_point(alpha = 0.8) +
  guides(size=FALSE) +
  theme(plot.title = element_blank(), legend.title = element_blank()) +
  coord_cartesian(ylim = c(30, 85)) +
  xlab("Fertility rate (births per woman)") +
  ylab("Life Expectancy") +
  geom_text(aes(x=7, y=82, label=year), cex=12, color="grey") +
  facet_grid(. ~ year) +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(),
        strip.text.y = element_blank(),
   legend.position = "top")
```


# Optional (Data Distributions & Outliers)

## Detecting outliers
For this part we use a different and more simple dataset
This dataset contains 1192 observations on self-reported:

 - `height` (inch)
 - `earn` ($)
 - `sex` (gender)
 - `ed` (currently unannotated)
 - `age` (years)
 - `race` 
 
```{r}
heights_data <- read_csv(file = file.path(root,
                                          "data",
                                          "heights_outliers.csv"))

heights_data
```

## Data characteristics
We will focus on the variable `height` here
```{r}
summary_heights_data <- heights_data %>%
  group_by(sex, age) %>%
  summarise(mean_height = mean(height, na.rm = TRUE),
            min_height = min(height),
            max_height = max(height)) %>%
  arrange(desc(mean_height))

summary_heights_data[c(1:4),]
```

From the above summary we can conclude that there are two outliers (presumably entry errors).

**Calculate the height in meters for each outlier in the `Console`** 1 inch = 0,0254 meters

**Please discuss the solution with your neighbour** 

## Checking the frequency distribution
```{r}

heights_data %>%
  ggplot(aes(x = height)) +
  geom_histogram(aes(stat = "identity"), bins = 200)
```

This distribution looks odd. 
When you see a large x-axis with no data plotted on it, it usually means there is an outlier. If you look carefully, you will spot two outliers around 600

## Boxplots to detect outliers
```{r}
heights_data %>%
  ggplot(aes(y = height)) +
  geom_boxplot()
```

So apparantly there is one data point that is way off from the rest of the distribution. Let's remove this point, using `filter()` from the `{dplyr}` package like we did before on the `gapminder` dataset.
```{r}
heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(y = height)) +
  geom_boxplot()

## by sex

heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(y = height, x = sex)) +
  geom_boxplot()

```

## New frequency distribution
Now let's plot a new distribution plot, this time we plot density, leaving the outlier out
```{r}
heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(height)) +
  geom_freqpoly(aes(y = ..density..))

## by sex
heights_data %>%
  dplyr::filter(height < 100) %>%
  ggplot(aes(height)) +
  geom_freqpoly(aes(y = ..density.., colour = sex))

```


## Checking normality with a qqplot
```{r}
## a qqplot provides a visual aid to assess whether a distribution is approaching normality
source(file = file.path(root, "code", "ggqq.R"))
height_data_outlier_removed <- heights_data %>%
  dplyr::filter(height < 100)
  gg_qq(height_data_outlier_removed$height) 
  
## formal statistical proof
shapiro.test(height_data_outlier_removed$height)
```

all data -> reject hypothesis that the sample has a normal distribution

## Test individual distributions
```{r}
males <- height_data_outlier_removed %>%
  dplyr::filter(sex == "male")

females <- height_data_outlier_removed %>%
  dplyr::filter(sex == "female")

shapiro.test(males$height)
shapiro.test(females$height)
## add shapiro for each sex

## we can do the same for age
shapiro.test(males$age)
shapiro.test(females$age)
```
