---
title: "Introduction to {dplyr}"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Data Analysis Using R - `r Sys.Date()` </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## When working with data you must:

* Figure out what you want to do
* Describe those tasks in the form of a computer program
* Execute the program

## the {dplyr} package
The dplyr package makes these steps fast and easy:

* dplyr simplifies how you can think about common data manipulation tasks.
* Simple "verbs", functions that correspond to the most common data manipulation tasks, to help you translate those thoughts into code.
* It uses efficient data storage backends, so you spend less time waiting for the computer.

## Data bases can be connected to dplyr

 * Besides in-memory data frames
 * {dplyr} also connects to out-of-memory, remote databases. 
 * By translating your R code into the appropriate SQL
 * Allows you to work with both types of data using the same set of tools.

## Packages

```{r}
library(ggplot2)
library(dplyr)
```

## Data: `nycflights13::flights`

* Built in `nycflights13` data frame. 
* Contains all `r nrow(nycflights13::flights)` flights that departed from New York City in 2013. 
* Data comes from the US [Bureau of Transportation Statistics](http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0), and is documented in `?nycflights13`

## To install the package containing the data
```{r}
# install.packages("nycflights13")
library(nycflights13)
dim(flights)
head(flights)
flights <- nycflights13::flights

```

__dplyr can work with data frames as is, but if you're dealing with large data, it's worthwhile to convert them to a `tbl_df`: this is a wrapper around a data frame that won't accidentally print a lot of data to the screen.__

## Single table verbs

Dplyr aims to provide a function for each basic verb of data manipulation:

* `filter()` (and `slice()`)
* `arrange()`
* `select()` (and `rename()`)
* `distinct()`
* `mutate()` (and `transmute()`)
* `summarise()`
* `sample_n()` (and `sample_frac()`)


## Filter rows 

* `filter()` allows you to select a subset of rows in a data frame. 
* The first argument is the name of the data frame. 
* The second and subsequent arguments are the expressions that filter the data frame:
For example, we can select all flights on January 1st, 2013 with:
```{r}
## look at the variable names
names(flights)

```

## Factor levels (grouping variable)
```{r}
flights$year %>% unique()
flights$year <- as.factor(flights$year)
levels(flights$year)
```

## The actual filtering (January 1st)
"January 1st" equals to `flights$month == 1 & flights$day == 1`
````{r}
# ?dplyr::filter  calls the help file for this function
january_1st <- filter(flights, month == 1 & day == 1)
january_1st
```

## Multiple filters
Expanding the search and using `%>%`
```{r}
february_1st <- flights %>%
  filter(month == 2, day == 1, arr_delay <= 0) %>%
           arrange(desc(carrier))
head(february_1st)
february_1st$arr_delay

names(flights)
alway_delayed <- flights %>% 
  filter(arr_delay > 0) %>% 
  arrange(desc(carrier))
 

alway_delayed %>% select(carrier, arr_delay, day, month) %>%
  arrange(desc(arr_delay)) %>%
  mutate(arr_delay_in_hours = arr_delay/60 )

```

## Base R equivalent
Selecting rows and columns by logical indexing

**Remember: `data_frame[rows, columns]`**

```{r}
index_logical <- flights$month == 1 & flights$day == 1
head(index_logical)
## selecting the proper rows
flights[index_logical,]
```

## Logical indexing in one call
```{r, eval = FALSE}
flights[flights$month == 1 & flights$day == 1, ]
## notice where we've put the comma
```

## Selecting the first column and the first ten rows with the index
```{r}
flights[c(1:10),1]
```

## Filter booleans
 
* `filter()` works similarly to `subset()` except that you can give it any number of filtering conditions, which are joined together with `&` (not `&&` which is easy to do accidentally!).  
* You can also use other boolean operators. 
* You can use `|` as the `OR` boolean operator, and `!` as `NOT`:

```{r, eval = FALSE}
## Data from January and February only
jan_feb <- filter(flights, month == 1 | month == 2)

## Delayed at departure in January and February
(dep_delayed <- flights %>% 
  filter(month == 1 | month == 2) %>%
  filter(!dep_delay <= 0)) # shows all dep_delay (except <= 0)
```

## `slice`: To select rows by position, use `slice()`

```{r}
slice(flights, c(1,5,7,8,9))
```
What is the base-R equivalent?

## Arrange rows with `arrange()`

* `arrange()` works similarly to `filter()` 
* It reorders rows  
* It takes a data frame and a set of column names (or more complicated expressions) to order by. 
* If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns
* Equivalent to `sort` function in MS Office Excel

```{r}
arrange(flights, year, month, day)
```

## Use `desc()` to order a column in descending order:

```{r}
x <- arrange(flights, arr_delay)
head(x$arr_delay)
```

## Base R equivalent
`dplyr::arrange()` works the same way as `plyr::arrange()`. It's a straightforward wrapper around `order()` that requires less typing. The previous code is equivalent to:

```{r, eval = FALSE}
flights[order(flights$year, flights$month, flights$day), ] 
flights[order(flights$arr_delay, decreasing = TRUE), ] 
flights[order(-flights$arr_delay), ]
## again look at the position of the comma
```

## Select columns with `select()`

* To select the columns of a dataframe that are actually of interest to you. 
* `select()` allows you to rapidly zoom in on a data subset:

```{r}
everything_but_month <- flights %>%
  select(month:arr_delay, year)

# Select columns by name
select(flights, year, month, day)
# Select all columns between year and day (inclusive)
select(flights, dep_delay:air_time)
# Select all columns except those from year to day (exclusive)
select(flights, -(year:day))
names(flights)
```


## Additional 'select' helper functions
There are a number of helper functions you can use within `select()`, like '

- `starts_with()`
- `ends_with()`
- `matches()` 
- `contains()`. 
These let you quickly match larger blocks of variables that meet some criterion. See `?select` for more details.

## Renaming variables
You can rename variables with `select()` by using named arguments:

```{r}
rename(flights, tail_num = tailnum)
```

## Extract distinct (unique) rows

Use `distinct()` to find unique values in a table:

```{r}
#unique(flights$carrier) # base-R
distinct(flights, carrier) # dplyr
names(flights)

flights$carrier <- as.factor(flights$carrier)
levels(flights$carrier)

unique(flights$tailnum)
unique(flights$origin)
unique(flights$dest)
```

(This is very similar to `base::unique()` but should be much faster.)

## Add new variable with `mutate()`

* To add new columns that are functions (calculations) of existing columns. 
* This is the job of `mutate()`:

```{r}
flights_mutated <- flights %>%
  mutate(gain = arr_delay - dep_delay,
  speed = distance / air_time * 60) %>%
  filter(gain <= 0) %>% 
  select(gain, speed, carrier, origin, month, day) %>%
  arrange(gain)

flights_mutated
```


## Another `mutate` example
```{r}
mutate(flights,
  gain = arr_delay - dep_delay,
  gain_per_hour = gain / (air_time / 60)
) 
```


## Summarise values with `summarise()`

The last verb is `summarise()`. It collapses a data frame to a single row (this is exactly equivalent to `plyr::summarise()`):

```{r}
summarise(flights,
  delay = mean(dep_delay, na.rm = TRUE))
names(flights)
summary <- flights %>% 
  group_by(carrier) %>%
  summarise(mean_air_time = mean(air_time, na.rm = TRUE)) %>%
  arrange(desc(mean_air_time))
summary  


```

## applying `summarise()`
Below, we'll see how this verb can be very useful.

## Randomly sample rows with `sample_n()` and `sample_frac()`

You can use `sample_n()` and `sample_frac()` to take a random sample of rows: use `sample_n()` for a fixed number  and `sample_frac()` for a fixed fraction.

```{r}
sample_n(flights, 10)
sample_frac(flights, 0.01)
```
## Commonalities

You may have noticed that the syntax and function of all these verbs are very similar:

* The first argument is a data frame.
* The subsequent arguments describe what to do with the data frame. Notice that you can referto columns in the data frame directly without using `$`.
* The result is a new data frame

Together these properties make it easy to chain together multiple simple steps to achieve a complex result.

## Grouped operations

Grouping affects the verbs as follows:

* grouped `select()` is the same as ungrouped `select()`, except that 
  grouping variables are always retained. 
* grouped `arrange()` orders first by the grouping variables

* `mutate()` and `filter()` are most useful in conjunction with window 
  functions (like `rank()`, or `min(x) == x`). They are described in detail in 
  `vignette("window-functions")`.
  
* `sample_n()` and `sample_frac()` sample the specified number/fraction of
  rows in each group.
  
* `slice()` extracts rows within each group.
  
* `summarise()` is powerful and easy to understand, as described in 
  more detail below.

In the following example, we split the complete dataset into individual planes and then summarise each plane by counting the number of flights (`count = n()`) and computing the average distance (`dist = mean(Distance, na.rm = TRUE)`) and arrival delay (`delay = mean(ArrDelay, na.rm = TRUE)`). We then use ggplot2 to display the output.

```{r, warning = FALSE, message = FALSE, fig.width = 6}
by_tailnum <- flights %>%
   group_by(tailnum) %>%
   summarise(count = n(),
   dist = mean(distance, na.rm = TRUE),
   delay = mean(arr_delay, na.rm = TRUE)) %>%
   arrange(desc(delay))
by_tailnum

(flights_slow <- flights %>%
  filter(tailnum == "D942DN"))


(delay <- filter(by_tailnum, count > 20, dist < 2000))
```

## Plotting
```{r} 
# Interestingly, the average delay is only slightly related to the
# average distance flown by a plane.
p <- ggplot(data = delay, aes(x = dist, y= delay)) +
  geom_point() +
  geom_smooth() +
  scale_size_area() +
  facet_wrap(~ count)
p
?geom_smooth
```

## Summarise with aggregate
You use `summarise()` with __aggregate functions__, which take a vector of values and return a single number. There are many useful examples of such functions in base R like `min()`, `max()`, `mean()`, `sum()`, `sd()`, `median()`, and `IQR()`. dplyr provides a handful of others:

* `n()`: the number of observations in the current group

* `n_distinct(x)`:the number of unique values in `x`.

* `first(x)`, `last(x)` and `nth(x, n)` - these work
  similarly to `x[1]`, `x[length(x)]`, and `x[n]` but give you more control
  over the result if the value is missing.

## Destinations
For example, we could use these to find the number of planes and the number of flights that go to each possible destination:

```{r}
destinations <- group_by(flights, dest)
summarise(destinations,
  planes = n_distinct(tailnum),
  flights = n()
)
```

## Inceremental summary (roll-up)
When you group by multiple variables, each summary peels off one level of the grouping. That makes it easy to progressively roll-up a dataset:

```{r}
daily <- group_by(flights, year, month, day)
(per_day   <- summarise(daily, flights = n()))
(per_month <- summarise(per_day, flights = sum(flights)))
(per_year  <- summarise(per_month, flights = sum(flights)))
```

However you need to be careful when progressively rolling up summaries like this: it's ok for sums and counts, but you need to think about weighting for means and variances (it's not possible to do this exactly for medians).

## Chaining
The dplyr API is functional in the sense that function calls don't have side-effects. You must always save their results. This doesn't lead to particularly elegant code, especially if you want to do many operations at once. You either have to do it step-by-step:

```{r, eval = FALSE}
a1 <- group_by(flights, year, month, day)
a2 <- select(a1, arr_delay, dep_delay)
a3 <- summarise(a2,
  arr = mean(arr_delay, na.rm = TRUE),
  dep = mean(dep_delay, na.rm = TRUE))
a4 <- filter(a3, arr > 30 | dep > 30)
```

## The complicated way
Or if you don't want to save the intermediate results, you need to wrap the function calls inside each other:

```{r}
filter(
  summarise(
    select(
      group_by(flights, year, month, day),
      arr_delay, dep_delay
    ),
    arr = mean(arr_delay, na.rm = TRUE),
    dep = mean(dep_delay, na.rm = TRUE)
  ),
  arr > 30 | dep > 30
)
```

## The pipe alternative
This is difficult to read because the order of the operations is from inside to out. Thus, the arguments are a long way away from the function. To get around this problem, dplyr provides the `%>%` operator. `x %>% f(y)` turns into `f(x, y)` so you can use it to rewrite multiple operations that you can read left-to-right, top-to-bottom:

```{r, eval = FALSE}
flights %>%
  group_by(year, month, day) %>%
  select(arr_delay, dep_delay) %>%
  summarise(
    arr = mean(arr_delay, na.rm = TRUE),
    dep = mean(dep_delay, na.rm = TRUE)
  ) %>%
  filter(arr > 30 | dep > 30)
```

## Other data sources

As well as data frames, dplyr works with data that is stored in other ways, like data tables, databases and multidimensional arrays.

