---
title: "Exercise - Vectors, Access, Select & Ordering"
author: "Marc A.T. Teunis"
date: '`r Sys.Date()`'
output:
  html_document:
    css: ~/hp_r_course/exercises/exercises_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      results = 'hide',
                      fig.show = 'hide',
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

```{r, packages, echo=TRUE}
library(tidyverse)
```

__Write an Rmd file, containing code chunks (if applicable) and answer all the questions below. Store the file in a folder called: "./answers_exercises" in your course project folder. Give this file the same name as the title of the current Rmd file__

# Vectors
 
## Character Vectors

1) Create a character vector named char that has capital letters "A", "B
```{r}
char <- rep(LETTERS[1:4], 4)
```

2) Select all instances of capital letter "B"
```{r}
ind <- char == "B"
ind
char[ind]

```

3) Replace all instances of capital letter "D" with "d"
```{r}
char[char == "D"] <- c("d")
char
```

4) Assign the letter "B" to a variable and then replace all instances of capital letter "B" in the vector character (char) with "b" using this variable. (Hint: use `tolower()`)
```{r}
#char <- rep(LETTERS[1:4], 4)
char
mychar <- "B"
char[char == mychar] <- tolower(mychar)
mychar
char
```

## Numeric vectors 

5) Create a numerical vector named numbers using the `runif()` function consisting of 30 values between -2 and 10.
```{r}
set.seed(10)
numbers <- runif(30, min=-2, max=10)
```

6) Select all values greater than 3. How many do you get? (Hint: use length) What is the sum of these values?
```{r}
length(numbers[numbers > 3])
sum(numbers[numbers > 3])
```

7) Select all values smaller than -1.2 or greater than 4. How many do you get?
```{r}
# step 1:`
positions <- numbers < -1.2 | numbers > 4
positions

#step 2:
length(numbers[positions])

# or in one go:
length(numbers[numbers < -1.2 | numbers > 4])
```

8) Assign missing values randomly to this vector using the following R code: `numbers[sample(1:30, sample(5:10, 1))] <- NA` . How many missing values do you have?
```{r}
numbers
numbers[sample(1:30, sample(5:10, 1))] <- NA
numbers

length(numbers[is.na(numbers)]) # or: 
sum(is.na(numbers))
# since sum(logical.index) converts the logical to 0 and 1 first!
```

# Missing values

9) Rewrite the statement from exercise 8 to also exclude missing values. (Hint: and operators have precedence over or operators) How many values do you get? What is the average of these values?
```{r}
# First filter:
numbers2 <- numbers[ !is.na(numbers) ] # after this, same as exercise 8
# OR:
sum( numbers <  -1.2 | numbers > 4, na.rm=TRUE)    # see Exercise 9
mean( numbers[ numbers <  -1.2 | numbers > 4 ], na.rm=TRUE )  # (version 23 nov contained error)
# OR: rely on fact that  NA & FALSE is still FALSE (and also NA | TRUE is TRUE)
ind <- (numbers < -1.2 | numbers > 4) & !is.na(numbers)
length(numbers[ind])
mean(numbers[ind])
```

