---
title: "Demo - Data Import; {readxl}"
author: "Marc A.T. Teunis"
date: '`r Sys.Date()`'
output:
  ioslides_presentation: 
    highlight: pygments
    logo: ~/hp_r_course/images/Logo INT_GOOD VERSION_VECTORS.png
    widescreen: yes
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Data Analysis Using R - July 2017 </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      results = 'markup')
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Lesson Aim
```{r, echo=FALSE, fig.align='right'}
path_to_readxl_sticker <- file.path(root, "images", "readxl_sticker.png")
 knitr::include_graphics(path = path_to_readxl_sticker, dpi = 350)
```

* Learn about importing Microsoft Office Excel files
* Excel is still widely used
* Most people that do knot know of "tidy" data provide "messy" data
* Even people that do know "tidy" data do not always provide "tidy" data

## The Excel Enigma

 * Excel is not meant for data wrangling
 * It is not really good for visualizations either
 * It is meant for rectangular 'tidy' data, but nobody does it!
 
## Reading "\*.xls" and "\*.xlsx" files with {readxl}
 
  * There are many R packages that can read Excel files
  * {readxl} does not need any external dependencies
  * {readxl} applies the {tidyverse} logic
  
## Real-life example
```{r, echo=FALSE, fig.align='center'}
path_to_messy_excel <- file.path(root, "images", "messy_excel.png")
knitr::include_graphics(path_to_messy_excel, dpi = 110)
```

## Let's read this messy Excel file into R
```{r}
path_to_messy_excel_data <- file.path(root, "data", "messy_excel.xlsx")
messy <- readxl::read_excel(path = path_to_messy_excel_data, sheet = 1 )
head(messy, 6)
```

## So that didn't work, do you know why?
 
 * There is stuff at the top that we do not need
 * There is stuff at the bottom that we do not want 
 * The dataset contains different variables, which?
 
## Variables in the dataset:
 
 * Laboratory
 * EC50
 * IL-1-alpha_2x
 * EC3
 * Compound
 
 How do we extract these and make the data tidy?
 
## Splitting the set into two: 
 
`EC50` and `IL-1alpha_2x` are side-by-side 

```{r} 
library(readxl)
library(tidyverse)
EC50 <- readxl::read_excel(path = path_to_messy_excel_data, 
                           sheet = 1,
                           range = cell_cols("C:F"),
                           na = c("not tested"),
                           col_names = c("hu", "disfeb",                                                "vumc", "basf")
)

head(EC50, 8)
```

## IL-1 alpha

```{r}
IL1alpha <- readxl::read_excel(path = path_to_messy_excel_data, 
                           sheet = 1,
                           range = cell_cols("I:L"),
                           na = c("not tested", "not reached"),
                           col_names = c("hu", "disfeb",                                                "vumc", "basf")
)
head(IL1alpha)
```

## Cleaning up the data further

You will notice that there are a number of stranges records in the data.
Extract the actual data with indexing

```{r}
EC50 <- EC50[c(3:15), ]
IL1alpha <- IL1alpha[c(3:15), ]
```

## The clean EC50 and IL1alpha data
```{r}
head(EC50, 3)
head(IL1alpha, 3)
```

## Adding an extra variable with compound names

```{r}
chemical_names <- readxl::read_excel(path_to_messy_excel_data,
                                     range = cell_cols("A"))

chemical_names <- chemical_names[c(4:16), ]
chemical_names <- chemical_names$`Correlation LLNA-EC3 data` 


EC50$chemical_names <- chemical_names
IL1alpha$chemical_names <- chemical_names
```

## Adding a new variable on the `read_out` variable

```{r}
EC50$read_out <- c("ec50")
IL1alpha$read_out <- c("il_1_alpha")
```

## Combining the two data
We can combine the data now. 

```{r}
library(tidyr)
tidy <- dplyr::bind_rows(EC50, IL1alpha)
```

## Complete Cases

```{r}
tidy <- tidy[complete.cases(tidy), ]

```

## Gather all the nummeric variables into one

The variables called `hu` to ` basf` are actually no variables
We can `gather()` these variables into onw called `laboratory` and put the numeric value in `result`

```{r}
library(tidyr)
tidy <- tidy %>%
  gather(hu:basf, key = laboratory, value = result)
```

## Checking variable class
```{r}
str(tidy)
tidy$chemical_names <- as.factor(tidy$chemical_names)
tidy$read_out <- as.factor(tidy$read_out)
tidy$laboratory <- as.factor(tidy$laboratory)
tidy$result <- as.numeric(tidy$result)
```

## Rounding digits
```{r}
tidy$result <- round(tidy$result, 2)
```

## Final result
```{r}
tidy
```

## Plotting the result 

We will make one plot to start the Exploratory Data Analysis.

Which chemical do you think is the most potent?

```{r}
plot <- tidy %>% 
  #filter(read_out == "ec50") %>%
ggplot(aes(x = reorder(chemical_names, 
                       result, FUN= mean), 
           y = result)) +
  geom_point() + 
  facet_grid(read_out ~ laboratory) + 
  coord_flip() +
  ylab("Result (mg/ml") +
  xlab("Chemical names") +
  ggtitle("Per lab and per chemical")
```

## Plot
```{r}
plot
```



