---
title: "dplyr - Exercise"
author: "Marc A.T. Teunis"
date: '`r Sys.Date()`'
output:
  html_document:
    css: ~/hp_r_course/exercises/exercises_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results = 'hide',
                      fig.show = 'hide',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE)
```

```{r, root_1, echo=TRUE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

__Write an Rmd file, containing code chunks (if applicable) and answer all the questions below. Store the file in a folder called: "./answers_exercises" in your course project folder. Give this file the same name as the title of the current Rmd file__

## The dplyr package is great for manipulating data

 * Select variables (columns)
 * Filter observaions (rows)
 * Change the order of rowns and columns 
 * Add new variables 
 * Group data / slice or "roll-up" data
 * Subset data
 * Summarize data
 
## dplyr has a few main "verbs"
 
 * `select()` to select columns/variables
 * `filter()` to select rows/observations
 * `mutate()` to create new variables
 * `group_by` to group that data
 * `summarize()` to calculate summary values (e.g. `mean()` or `sd()`)
 * `arrange()` to sort observations. If you want descending use `desc()` within the `arrange()` call like `arrange(desc(variable_for_which_to_sort))` 
 * `count()` to count items in a vector or column
 
## The exercise data

For this exercise we use the "heights" dataset.
This dataset contains heights, weights, income and ethnicity information. The information for the variables that are needed for this exercise are in the file "annotations_height.txt".

It was downloaded from:

https://github.com/rstudio/webinars/tree/master/15-RStudio-essentials/4-Projects

Load the data with the following code
```{r, echo=TRUE}
library(tidyverse)
heights_file <- file.path(root, "data", "heights.RDS")
heights <- read_rds(path = heights_file)
head(heights)
heights
heights_tib <- as_tibble(heights)

```

## Data inspection

1A) Inspect the data with the `head()` and `tail()` function.
The datset contains `r nrow(heights)` observations.   

 - What types of variables are present in the data.
 - Focus on the variables `id` `income`, `height`, `weight`, `sex` and `race`
 
1B) Call `summary` on the data. What do you notice when you look at the max. and min. value for the `height`. For nicer printing you can use `pander::pander()`

```{r}
library(pander)
pander(summary(heights))
```

1C) Are there any missing values in the data? Determine the number of missing values
```{r}
sum(is.na(heights))
```
 
## Selecting variables with {dplyr} and missing values

Subsetting data with the {dplyr} verbs `select()` and `filter()` is convenient for creating slices of the data that are of interest.

For this exercise we will use the variables: 

`id`, `income`, `height`, `weight`, `sex` and `race`

2A) Generate a subset of the `heights` dataframe containing only these variables. Call this new dataframe `heights_selected`
 
 - Use `dplyr::select()`
 - Use `%>%` 
 - To find NAs, use `sum(is.na(dataframe))`

```{r, answer_2a}
heights_selected <- heights %>%
  select(id, income, height, weight, sex, race)
```

2B) Check the amount of NAs that you have left in the dataset
```{r, answer_2b}
sum(is.na(heights_selected))
```

2C) To complete the data-inspection and do a bit of cleaning we can look at the records in the `heigths_selected` dataset that are complete. This means, for which we have data points for all 5 variables.

use the function `complete.cases()` from the {stats} package. Look at `?complete.cases()` to see how to use this function. Create a new data frame with complete case that you call `heights_complete`

 - Remember that you can use data.frame[rows, columns] to subset a dataframe with a logical vector
 
```{r}
library(stats)
complete_cases <- complete.cases(heights_selected) 
heights_complete <- heights_selected[complete_cases,] 

## or


```

2D) How many NAs are present in the `heights_complete` dataset that you just created under 2C? What did you expect?
```{r}
sum(is.na(heights_complete))
```

## Filtering

 - For the next assignment you can use the `filter()` command from {dplyr}

3A) Filter all people that have a weight of over 200 lbs.
```{r}
heights_complete <- tibble(heights_complete)

over_200 <- heights_complete %>% 
 filter(weight > 200)

over_200 <- filter(heights_complete, weight > 200)

```

3B) What is the most frequent ethnicity in the group of people that have a weight over 200 lbs   

 - Think about how you can answer this question: you want to have some sort of tabular summary! Discuss with your neighbour. And/or look on the internet. 
```{r}

haviest_race <- over_200 %>%
    summarise(frequency = count(weight)) %>%
  arrange(desc(mean_weight))
haviest_race

summary <- summary(over_200)
summary
str(summary)
table(summary[, 6])


# so the answer is "other", which is not surprising
```

3C) Filter to extract the top 5 tallest hispanic males in the dataset.  
What is their respective weight? 
What is their average height?

 - This one is already a bit more complicated:
 - Use `%>%`
 - Use `filter()`
 - Use `arrange()` to sort in descending order

```{r}
# filter for 'hispanic' 'males'
tallest_hispanic_men <- heights_complete %>%
  filter(race == "hispanic" & sex == "male") %>%
# now sort the rows in descending order
  arrange(desc(height))
# select only the first five rows
five_tallest_hispanic_men <- tallest_hispanic_men[c(1:5), ]

# all weights: 
five_tallest_hispanic_men$height


# mean height
mean_height_five_tallest_hispanic_men <- 
mean_height_five_tallest_hispanic_men

  mean(five_tallest_hispanic_men$height)

  mean_height_five_tallest_hispanic_men

```

3D) Find the two tallest women in the data. See if you can find two men that match the weight and heigt of these women.
```{r}
rank_women <- heights_complete %>%
  filter(sex == "female") %>%
  arrange(desc(height))

(two_tallest_women <- rank_women[c(1,2), ])


match_1 <- heights_complete %>%
  filter(sex == "male" & 
         weight == two_tallest_women$weight[1] & 
         height == two_tallest_women$height[1])

match_1

match_2 <- heights_complete %>%
  filter(sex == "male" & 
         weight == two_tallest_women$weight[2] &
         height == two_tallest_women$height[2])
match_2
```

## Creating new variables
The units of the data in the `heights` dataset is somewhat difficult to interpret because of the use of non-SI notation (so basically non-scientific units). The weights are in `lbs` or also called pounds and the height is in `inches`.

 * 1 inch = 2.54 cm
 * 1 lbs = 0.45359237 kg

4A) Convert the `height` and the `weight` variable to SI units (so meters, and kg, respectively)
 
 Add the new variables to the exsting `heigths_complete` dataset. Call the new columns: `height_m` and `weight_kg`
 
  - Use `dplyr::mutate()` to solve this question 
  - Round the `height_m` and `weight_kg` to decimals that make sense
 
```{r}
heights_complete <- heights_complete %>%
  mutate(height_m = (height*2.54/100),  
         weight_kg = weight*0.45359237)

heights_complete


heights_complete$height_m <- round(heights_complete$height_m, 2)
heights_complete$weight_kg <- round(heights_complete$weight_kg, 1)

head(heights_complete) 
```

## Body-mass index
The body-mass index or BMI can be calculated from the weight and the height of a person. 
 
 The formula for determining the BMI is
 
 Bodyweight in kilograms divided by height in meters squared
 
 or:
 
#### $BMI=\frac{bodyweight(kg)}{height(meters)^2}$
 
4C) Calculate a new variale called `bmi` for the `heights_complete` dataset 
```{r}
heights_complete <- heights_complete %>%
  mutate(bmi = (weight_kg / height_m^2))
head(heights_complete)
 
```

## Grouping and summarizing

```{r}
names(heights_complete)


heights_summary <- heights_complete %>%
  select(-id) %>%
  group_by(race, sex) %>%
  summarise(mean_height_in = mean(height), 
            mean_weight_lbs = mean(weight), 
            mean_height_m = mean(height_m),
            mean_weight_kg = mean(weight_kg), 
            mean_bmi = mean(bmi), 
            mean_income = mean(income)) %>%
  arrange(desc(mean_income))

heights_summary[1, "race"]

head(heights_summary)

```




When we have multiple observations per unit or groups of units, it makes sense to be able to group the data accordingly.

Here we need to take a closer look at `dplyr::group_by` and `dplyr::summarize` to group, summarize and generate new summary variables of the (cleaned) data. 

5A) Summarize the data.

 - Group the data by `race`, and `sex`, drop the `id`, variable
 - Call the new summarized dataframe: `heights_summary`
 - Make a `summary` dataframe with new summary variables with the `dplyr::summary()`function 
 - Call the new summary variables `mean_height_in`, `mean_weight_lbs`, `mean_height_m`, `mean_weight_kg`, `mean_bmi` and `mean_income`
 - Rank (`dplyr::arrange()`) the data according income from high to low (`arrange(desc(variable_for_which_to_rank)`)
 
```{r} 
names(heights_complete)
heights_summary <- heights_complete %>%
  select(income:bmi) %>%
  group_by(race, sex) %>%
  summarise(mean_height_in = mean(height),
            mean_height_m = mean(height_m),
            mean_weight_lbs = mean(weight),
            mean_weight_kg = mean(weight_kg),
            mean_bmi = mean(bmi),
            mean_income = mean(income)) %>%
  arrange(desc(mean_income))


```

5B) Which race has the highest income?
 
```{r} 
names(heights_complete)
heights_summary <- heights_complete %>%
  select(income:bmi) %>%
  group_by(race, sex) %>%
  summarise(mean_height_in = mean(height),
            mean_height_m = mean(height_m),
            mean_weight_lbs = mean(weight),
            mean_weight_kg = mean(weight_kg),
            mean_income = mean(income)) %>% 
  arrange(desc(mean_income)) 
```

5C) Is there a difference between males and females. For which race is this difference the biggest?

```{r}
males <- heights_summary %>%
  filter(sex == "male") %>% 
  select(mean_income, race) %>%
  arrange(desc(race))

females <- heights_summary %>%
  filter(sex == "female") %>%
  select(mean_income, race) %>%
  arrange(desc(race))

difference <- as_tibble(cbind(males, females)) 
difference

names(difference) <- c("mean_income_m", "race_m", "mean_income_f", "race_f")
names(difference)
# str(difference)

difference %>%
  mutate(delta_income = (mean_income_m - mean_income_f)) #%>%
  #  arrange(delta_income)
```

