---
title: "Data Analytic Using R - Contents"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---
<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Helvetica'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Analytics Using R - `r Sys.Date()` </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Programme

 * Use Case: "Open Analytics" 30 min.
 * Demo / Hands-on 1; Github & RStudio
 
 ------ break ------
 
 * Demo / Hands-on 2; `{rmarkdown}` for reproducibility
 * Demo / Hands-on 3; Visualizations with `{ggplot2}`
 
 ------ break ------
 
 * Use case "......" 
 * Demo / Hands-on 4; Data wrangling with `{dplyr}`
 * (Reserve: Hands-on 5; Building R packages




